---
title: "Ames Housing Dataset Regression Analysis"
author: "Lucas McGill"
date: today
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    theme: darkly
mainfont: "Georgia"
sansfont: "Helvetica Neue"
monofont: "JetBrains Mono"
---

## Introduction

This analysis explores the Ames Housing dataset to build a regression model predicting home sale prices. The dataset contains 2,930 residential property sales in Ames, Iowa from 2006-2010, with 82 variables describing various property characteristics.

**Objective:** Develop a multiple linear regression model to predict `SalePrice` using relevant predictor variables, demonstrating iterative model improvement and proper documentation of the prompt engineering process with a Large Language Model (LLM).

---

## Step One: Description of Data

### Data Loading and Initial Exploration

*Prompt 10: Load the Ames Housing dataset and show me the structure and dimensions of the data.*

```{r}
# Load the housing data
housing <- read.csv("amesHousing2011.csv")

# View first few rows
head(housing)

# Check dimensions
cat("Dataset dimensions:", nrow(housing), "rows and", ncol(housing), "columns\n")

# View structure of the dataset
str(housing)
```

**R Code Explanation:**
- `read.csv()` loads the CSV file into a data frame called `housing`
- `head()` displays the first 6 rows to see what the data looks like
- `nrow()` and `ncol()` show the number of rows (observations) and columns (variables)
- `str()` shows the structure: variable names, types (numeric, character, etc.), and first few values

### Summary Statistics

*Prompt 11: Generate summary statistics for all variables in the dataset.*

```{r}
# Summary statistics for all variables
summary(housing)
```

**R Code Explanation:**
- `summary()` provides descriptive statistics for each variable
- For numeric variables: min, 1st quartile, median, mean, 3rd quartile, max
- For character/factor variables: counts of each category
- Also shows count of NA's (missing values) if present

### Exploring the Response Variable: Sale Price

*Prompt 12: Show me the distribution of SalePrice with a histogram and boxplot to identify outliers.*

```{r}
# Histogram of Sale Price
hist(housing$SalePrice,
     breaks = 30,
     main = "Distribution of Sale Prices",
     xlab = "Sale Price ($)",
     col = "lightblue",
     border = "black")

# Summary statistics for SalePrice
cat("\nSale Price Summary:\n")
summary(housing$SalePrice)
cat("\nStandard Deviation:", sd(housing$SalePrice), "\n")

# Boxplot to identify outliers
boxplot(housing$SalePrice,
        main = "Boxplot of Sale Prices",
        ylab = "Sale Price ($)",
        col = "lightgreen")

# Identify potential outliers (above 95th percentile)
price_95 <- quantile(housing$SalePrice, 0.95)
cat("\n95th percentile of Sale Price: $", price_95, "\n")
cat("Number of sales above 95th percentile:",
    sum(housing$SalePrice > price_95), "\n")
```

**R Code Explanation:**
- `hist()` creates a histogram showing the distribution of sale prices
  - `breaks = 30` creates 30 bins for more detail
  - Right-skewed distribution expected (few expensive homes)
- `boxplot()` visualizes outliers as points beyond the whiskers
- `quantile()` calculates percentiles to identify outlier threshold
- We identify homes above the 95th percentile as potential outliers (mansions)

**Observations:**
- The distribution appears right-skewed with some high-value outliers
- These outliers (mansion sales) may distort our regression model
- Professor recommendation: Remove these outliers before modeling

### Exploring Key Predictor Variables

*Prompt 13: Show me distributions of the most important predictor variables: living area, overall quality, and year built.*

```{r}
# Living Area (GrLivArea)
par(mfrow = c(2, 2))  # Create 2x2 plot layout

hist(housing$GrLivArea,
     breaks = 30,
     main = "Distribution of Living Area",
     xlab = "Above Ground Living Area (sq ft)",
     col = "lightcoral")

# Overall Quality
hist(housing$OverallQual,
     breaks = 10,
     main = "Distribution of Overall Quality",
     xlab = "Overall Quality (1-10)",
     col = "lightyellow")

# Year Built
hist(housing$YearBuilt,
     breaks = 30,
     main = "Distribution of Year Built",
     xlab = "Year Built",
     col = "lightcyan")

# Total Basement Square Feet
hist(housing$TotalBsmtSF,
     breaks = 30,
     main = "Distribution of Basement Area",
     xlab = "Total Basement SF",
     col = "lavender")

par(mfrow = c(1, 1))  # Reset to single plot

# Summary statistics for key predictors
cat("\nLiving Area Summary:\n")
summary(housing$GrLivArea)

cat("\nOverall Quality Summary:\n")
summary(housing$OverallQual)

cat("\nYear Built Summary:\n")
summary(housing$YearBuilt)
```

**R Code Explanation:**
- `par(mfrow = c(2, 2))` sets up a 2×2 grid to display 4 plots together
- Each histogram shows the distribution of a key predictor variable
- `par(mfrow = c(1, 1))` resets the plot layout back to single plots
- Summary statistics help identify the range and central tendency

### Relationships Between Variables

*Prompt 14: Create scatterplots showing the relationship between SalePrice and key predictors.*

```{r}
# Set up 2x2 plot layout
par(mfrow = c(2, 2))

# SalePrice vs Living Area
plot(housing$GrLivArea, housing$SalePrice,
     main = "Sale Price vs Living Area",
     xlab = "Living Area (sq ft)",
     ylab = "Sale Price ($)",
     pch = 19,
     col = rgb(0, 0, 1, 0.3))  # Semi-transparent blue

# SalePrice vs Overall Quality
plot(housing$OverallQual, housing$SalePrice,
     main = "Sale Price vs Overall Quality",
     xlab = "Overall Quality (1-10)",
     ylab = "Sale Price ($)",
     pch = 19,
     col = rgb(1, 0, 0, 0.3))  # Semi-transparent red

# SalePrice vs Year Built
plot(housing$YearBuilt, housing$SalePrice,
     main = "Sale Price vs Year Built",
     xlab = "Year Built",
     ylab = "Sale Price ($)",
     pch = 19,
     col = rgb(0, 1, 0, 0.3))  # Semi-transparent green

# SalePrice vs Basement Area
plot(housing$TotalBsmtSF, housing$SalePrice,
     main = "Sale Price vs Basement Area",
     xlab = "Total Basement SF",
     ylab = "Sale Price ($)",
     pch = 19,
     col = rgb(0.5, 0, 0.5, 0.3))  # Semi-transparent purple

par(mfrow = c(1, 1))  # Reset plot layout
```

**R Code Explanation:**
- `plot(x, y)` creates scatterplots to visualize relationships
- `pch = 19` uses filled circles as plot points
- `rgb(r, g, b, alpha)` creates semi-transparent colors to see overlapping points
- We're looking for:
  - **Linear relationships**: Points form a straight line pattern
  - **Non-linear relationships**: Curved patterns (may need transformation)
  - **Heteroscedasticity**: Spread of points increases with X (funneling)
  - **Strength**: How tightly points cluster around a pattern

**Expected Observations:**
- **Living Area**: Strong positive linear relationship expected
- **Overall Quality**: Positive relationship, possibly non-linear (may need log transformation)
- **Year Built**: Positive relationship, variance may increase with year (heteroscedasticity)

### Correlation Analysis

*Prompt 15: Calculate correlations between SalePrice and numeric predictors to identify the strongest relationships.*

```{r}
# Select key numeric variables
numeric_vars <- c("SalePrice", "GrLivArea", "OverallQual", "OverallCond",
                 "YearBuilt", "YearRemod.Add", "TotalBsmtSF", "X1stFlrSF",
                  "X2ndFlrSF", "BedroomAbvGr", "TotRmsAbvGrd",
                  "GarageCars", "GarageArea")

# Calculate correlation matrix
cor_matrix <- cor(housing[, numeric_vars], use = "complete.obs")

# Display correlations with SalePrice, sorted by absolute value
cor_with_price <- cor_matrix[, "SalePrice"]
cor_sorted <- sort(abs(cor_with_price), decreasing = TRUE)

cat("Correlations with Sale Price (sorted by strength):\n")
print(round(cor_sorted, 3))

cat("\n\nTop 10 correlations with SalePrice:\n")
print(round(cor_with_price[names(cor_sorted)[1:10]], 3))
```

**R Code Explanation:**
- `cor()` calculates correlation coefficients between variables
- `use = "complete.obs"` handles missing data by using only complete cases
- Correlation ranges from -1 (perfect negative) to +1 (perfect positive)
- Values close to 0 indicate weak or no linear relationship
- `sort(abs(...))` sorts by absolute value to find strongest relationships regardless of direction

**Important Note from Professor:**
Pairwise correlations can miss important relationships. For example, location variables may have mediating effects on other predictors that won't show up in simple correlations.

### Missing Data Analysis

*Prompt 16: Identify which variables have missing data and how much.*

```{r}
# Count missing values per variable
missing_counts <- colSums(is.na(housing))

# Show only variables with missing data
vars_with_missing <- missing_counts[missing_counts > 0]

if (length(vars_with_missing) > 0) {
  cat("Variables with missing data:\n")

  # Calculate percentage missing
  missing_pct <- 100 * vars_with_missing / nrow(housing)

  # Create summary data frame
  missing_summary <- data.frame(
    Variable = names(vars_with_missing),
    Count = vars_with_missing,
    Percentage = round(missing_pct, 2)
  )

  # Sort by count
  missing_summary <- missing_summary[order(-missing_summary$Count), ]

  print(missing_summary, row.names = FALSE)
} else {
  cat("No missing data detected in the dataset.\n")
}

# Total missing values
cat("\nTotal missing values in dataset:", sum(is.na(housing)), "\n")
```

**R Code Explanation:**
- `is.na()` returns TRUE for missing values
- `colSums()` counts missing values for each column
- We calculate both counts and percentages to assess severity
- Variables with >20% missing may need to be excluded
- Variables with <5% missing can use simple imputation
- **Professor recommendation**: Use multiple imputation instead of simple median imputation

### Summary of Key Findings

Based on the initial exploration:

1. **Dataset**: 2,930 observations with 82 variables describing housing characteristics

2. **Response Variable (SalePrice)**:
   - Right-skewed distribution
   - Outliers present (mansion sales) - will need to remove these
   - Will likely benefit from log transformation

3. **Key Predictors Identified**:
   - Living area (GrLivArea) - likely strongest predictor
   - Overall quality (OverallQual) - may have non-linear relationship
   - Year built - shows heteroscedastic relationship
   - Location variables - important per real estate folklore

4. **Data Quality Issues**:
   - Missing data in several variables
   - Outliers in sale price need addressing
   - Some relationships appear non-linear

**Next Steps**: Proceed to Step Two - data preprocessing and regression modeling.

